Partitioned Local Depth (PaLD) Clustering Analyses in R

### 2022-03-17 r-journal/manuscript.tex
update paper

\hypertarget{helper-functions-to-create-contribution-matrix}{%
\subsubsection{Helper functions to create contribution
matrix}\label{helper-functions-to-create-contribution-matrix}}

@@

\begin{Schunk}
\begin{Sinput}
library(pald)
df <- data.frame(
  x1 = c(6, 8, 11, 16, 4),
  x2 = c(5, 4, 13, 7, 18)
)
rownames(df) <- c("A", "B", "C", "D", "E")
\end{Sinput}
\end{Schunk}

@@

This will create an \(n\times n\) distance matrix, where \(n\)
corresponds to the number of observations in the original data frame, in
this example \(n = 5\). This distance matrix can then be passed to the
\texttt{cohesion\_matrix()} function in order to calculate the pairwise
cohesion values. Cohesion is an interpretable probability that reflects
the strength of alignment of two points. Again, if the user begins with
a distance matrix, they can skip the first step and simply input the
distance matrix into this function.

\begin{Schunk}
\begin{Sinput}
d <- dist(df)
cohesion_matrix(d)
\end{Sinput}
\begin{Soutput}
#>        A      B         C         D         E
#> A 0.2875 0.1625 0.0000000 0.0500000 0.0000000
#> B 0.1625 0.2875 0.0000000 0.0500000 0.0000000
#> C 0.0000 0.0000 0.3083333 0.1833333 0.1000000
#> D 0.0500 0.0500 0.1750000 0.3000000 0.0000000
#> E 0.0000 0.0000 0.1000000 0.0000000 0.2333333
\end{Soutput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
df |>
  dist() |>
  cohesion_matrix()
\end{Sinput}
\begin{Soutput}
#>        A      B         C         D         E
#> A 0.2875 0.1625 0.0000000 0.0500000 0.0000000
#> B 0.1625 0.2875 0.0000000 0.0500000 0.0000000
#> C 0.0000 0.0000 0.3083333 0.1833333 0.1000000
#> D 0.0500 0.0500 0.1750000 0.3000000 0.0000000
#> E 0.0000 0.0000 0.1000000 0.0000000 0.2333333
\end{Soutput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
d |>
  dist() |>
  cohesion_matrix() -> cohesion
\end{Sinput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
community_clusters(cohesion)
\end{Sinput}
\begin{Soutput}
#>   point cluster
#> A     A       1
#> B     B       1
#> C     C       2
#> D     D       2
#> E     E       3
\end{Soutput}
\end{Schunk}

In this example, three clusters are identified with these five points.
Points \texttt{A} and \texttt{B} fall into cluster 1. Points \texttt{C}
and \texttt{D} into cluster 2, and point \texttt{E} in cluster 3.

@@

\begin{Schunk}
\begin{Sinput}
local_depths(cohesion)
\end{Sinput}
\begin{Soutput}
#>         A         B         C         D         E 
#> 0.4062500 0.5416667 0.3833333 0.4687500 0.7000000
\end{Soutput}
\end{Schunk}

In this case, the deepest point is \texttt{C}.

@@

\begin{Schunk}
\begin{Sinput}
strong_threshold(cohesion)
\end{Sinput}
\begin{Soutput}
#> [1] 0.14
\end{Soutput}
\end{Schunk}

In this case, the threshold is \texttt{0.14}.

@@

\begin{Schunk}
\begin{Sinput}
cohesion_strong(cohesion)
\end{Sinput}
\begin{Soutput}
#>        A      B     C     D      E
#> A 0.3000 0.1625 0.000 0.000 0.0000
#> B 0.1625 0.2875 0.000 0.000 0.0000
#> C 0.0000 0.0000 0.300 0.175 0.0000
#> D 0.0000 0.0000 0.175 0.300 0.0000
#> E 0.0000 0.0000 0.000 0.000 0.2125
\end{Soutput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
graphs <- community_graphs(cohesion)
graphs[["G_strong"]]
\end{Sinput}
\begin{Soutput}
#> IGRAPH 261f39e UNW- 5 2 -- 
#> + attr: name (v/c), weight (e/n)
#> + edges from 261f39e (vertex names):
#> [1] A--B C--D
\end{Soutput}
\end{Schunk}

Here we see that there are two connected components, points \texttt{A}
and \texttt{B}, which form the first cluster, and points \texttt{D} and
\texttt{C} which form the second.

@@

The final category of function is functions for data visualization. We
can begin by visualizing the points in data frame \texttt{df} (Figure
\ref{fig:fig1}).

\begin{Schunk}
\begin{Sinput}
library(ggplot2)
ggplot(df, aes(x1, x2)) +
  geom_text(label = rownames(df))
\end{Sinput}
\begin{figure}
\includegraphics{manuscript_files/figure-latex/fig1-1} \caption[Visualize the points from data frame `df`]{Visualize the points from data frame `df`}\label{fig:fig1}
\end{figure}
\end{Schunk}

We can then pass the cohesion matrix to the
\texttt{plot\_community\_graphs()} function to view the relationship
between points (Figure \ref{fig:fig2}).

\begin{Schunk}
\begin{Sinput}
df |>
  dist() |>
  cohesion_matrix() |>
  plot_community_graphs()
\end{Sinput}
\begin{figure}
\includegraphics{manuscript_files/figure-latex/fig2-1} \caption[PaLD graph displaying the relationship between the points in data frame `df`]{PaLD graph displaying the relationship between the points in data frame `df`}\label{fig:fig2}
\end{figure}
\end{Schunk}

The \texttt{layout} argument allows the user to pass a matrix to dictate
the layout of the graph. For example, if we wanted the graph to match
the visualization displayed in Figure \ref{fig:fig1}, we can pass
\texttt{as.matrix(df)}, or a matrix of the data frame \texttt{df} to the
\texttt{layout} argument (Figure \ref{fig:fig3}.

\begin{Schunk}
\begin{Sinput}
df |>
  dist() |>
  cohesion_matrix() |>
  plot_community_graphs(layout = as.matrix(df))
\end{Sinput}
\begin{figure}
\includegraphics{manuscript_files/figure-latex/fig3-1} \caption[PaLD graph displaying the relationship between the points in data frame `d`, matching the original layout in Figure 1]{PaLD graph displaying the relationship between the points in data frame `d`, matching the original layout in Figure 1}\label{fig:fig3}
\end{figure}
\end{Schunk}

This \texttt{plot\_community\_graphs()} function will also permit
parameters that can be passed to the \texttt{plot.igraph()} function.
For example, to add axes to the graph, the user can pass the
\texttt{axes\ =\ TRUE} argument to the \texttt{...} in the
\texttt{plot\_community\_graphs()} function (Figure \ref{fig:fig4}).

\begin{Schunk}
\begin{Sinput}
df |>
  dist() |>
  cohesion_matrix() |>
  plot_community_graphs(layout = as.matrix(df),
                        axes = TRUE)
\end{Sinput}
\begin{figure}
\includegraphics{manuscript_files/figure-latex/fig4-1} \caption[PaLD graph displaying the relationship between the points in data frame `d`, matching the original layout in Figure 1, adding axes]{PaLD graph displaying the relationship between the points in data frame `d`, matching the original layout in Figure 1, adding axes}\label{fig:fig4}
\end{figure}
\end{Schunk}

@@

The first example if from a subset of tissue gene expression data from
\citet{zilliox2007gene}, \citet{mccall2011gene}, and
\citet{mccall2014gene}, obtained from the \textbf{tissuesGeneExpression}
bioconductor package \citep{tissue}. A cohesion matrix was created using
this data set and is included the \CRANpkg{pald} package in an object
called \texttt{tissue\_cohesion\_matrix}.

The \texttt{tissue\_cohesion\_matrix} object is a cohesion matrix with
189 rows and 189 columns.

We can use this contribution matrix to display the relationship between
tissue samples using the \texttt{plot\_community\_graphs()} function
(Figure \ref{fig:fig5}). For clarity of the display, we show how to only
include a random set of the vertex labels. We can pass this random set
through the \texttt{...} to the \texttt{plot.igraph}
\texttt{vertex.label} parameter.

\begin{Schunk}
\begin{Sinput}
set.seed(1)

labels <- rownames(tissue_cohesion_matrix)
labels[sample(1:189, 125)] <- ""
plot_community_graphs(tissue_cohesion_matrix,
                      vertex.label = labels,
                      vertex.size = 4)
\end{Sinput}
\begin{figure}
\includegraphics{manuscript_files/figure-latex/fig5-1} \caption[PaLD clustering of tissue data]{PaLD clustering of tissue data}\label{fig:fig5}
\end{figure}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
library(dplyr)
community_clusters(tissue_cohesion_matrix) %>%
  group_by(cluster, point) %>%
  count()
\end{Sinput}
\begin{Soutput}
#> # A tibble: 19 x 3
#> # Groups:   cluster, point [19]
#>    cluster point           n
#>      <dbl> <chr>       <int>
#>  1       1 endometrium    15
#>  2       1 kidney         39
#>  3       2 hippocampus    31
#>  4       3 cerebellum     26
#>  5       4 cerebellum      1
#>  6       5 colon          33
#>  7       6 colon           1
#>  8       7 liver           7
#>  9       8 cerebellum      1
#> 10       9 liver          17
#> 11      10 cerebellum      2
#> 12      11 liver           2
#> 13      12 cerebellum      1
#> 14      13 cerebellum      4
#> 15      14 cerebellum      2
#> 16      15 cerebellum      1
#> 17      16 placenta        2
#> 18      17 placenta        1
#> 19      18 placenta        3
\end{Soutput}
\end{Schunk}

@@

@@

Here, we will demonstrate how to use \texttt{exdata3}. These points were
generated from bivariate normal distributions with varying means and
variances.

\begin{Schunk}
\begin{Sinput}
exdata_cohesion_matrix <- exdata3 |>
  dist() |>
  cohesion_matrix()
\end{Sinput}
\end{Schunk}

When plotting the \texttt{exdata3} PaLD graph, we want the layout to
match the layout of the original data, so we will pass
\texttt{as.matrix(exdata3)} to the \texttt{layout} parameter in the
\texttt{plot\_community\_graphs()} function. Additionally, here the row
names are meaningless, they just correspond to the location of the
generated data, so we can remove the labels on the plot by setting
\texttt{show\_labels\ =\ FALSE}. The \texttt{only\_strong} parameter
will only display the strongly connected edges. Additionally, we can
optionally pass parameters through the \texttt{...} to
\texttt{plot::igraph} function; for example, to increase the vertex size
we can set \texttt{vertex.size\ =\ 5}. Figure \ref{fig:fig6} shows this.

\begin{Schunk}
\begin{Sinput}
plot_community_graphs(exdata_cohesion_matrix,
                layout = as.matrix(exdata3),
                show_labels = FALSE, 
                only_strong = TRUE,
                vertex.size = 5)
\end{Sinput}
\begin{figure}
\includegraphics{manuscript_files/figure-latex/fig6-1} \caption[PaLD clustering of randomly generated example data (from Figure 4D from Berenhaut et al]{PaLD clustering of randomly generated example data (from Figure 4D from Berenhaut et al. (2022))}\label{fig:fig6}
\end{figure}
\end{Schunk}

>>>>

\hypertarget{creating-the-cohesion-matrix}{%
\subsubsection{Creating the cohesion
matrix}\label{creating-the-cohesion-matrix}}

@@

\begin{Schunk}
\begin{Sinput}
library(pald)
df <- data.frame(
  x1 = c(6, 8, 11, 16, 4, 14),
  x2 = c(5, 4, 13, 7, 6, 10)
)
rownames(df) <- c("A", "B", "C", "D", "E", "F")
\end{Sinput}
\end{Schunk}

@@

This will create a \texttt{dist} object. If converted to a matrix, this
will be a \(n\times n\) distance matrix, where \(n\) corresponds to the
number of observations in the original data frame, in this example
\(n = 6\).

This dist object, or equivalently a distance matrix, can then be passed
to the \texttt{cohesion\_matrix()} function in order to calculate the
pairwise cohesion values. Cohesion is an interpretable probability that
reflects the strength of alignment of two points. Again, if the user
begins with a distance matrix, they can skip the first step and simply
input the distance matrix into this function.

\begin{Schunk}
\begin{Sinput}
d <- dist(df)
cohesion_matrix(d)
\end{Sinput}
\begin{Soutput}
#>           A         B         C         D         E         F
#> A 0.2333333 0.1666667 0.0000000 0.0000000 0.1666667 0.0000000
#> B 0.1333333 0.2333333 0.0000000 0.0000000 0.1000000 0.0000000
#> C 0.0000000 0.0000000 0.2333333 0.1000000 0.0000000 0.1000000
#> D 0.0000000 0.0000000 0.1000000 0.2666667 0.0000000 0.1666667
#> E 0.1333333 0.1000000 0.0000000 0.0000000 0.2333333 0.0000000
#> F 0.0000000 0.0000000 0.1000000 0.1666667 0.0000000 0.2666667
#> attr(,"class")
#> [1] "cohesion_matrix" "matrix"          "array"
\end{Soutput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
df |>
  dist() |>
  cohesion_matrix()
\end{Sinput}
\begin{Soutput}
#>           A         B         C         D         E         F
#> A 0.2333333 0.1666667 0.0000000 0.0000000 0.1666667 0.0000000
#> B 0.1333333 0.2333333 0.0000000 0.0000000 0.1000000 0.0000000
#> C 0.0000000 0.0000000 0.2333333 0.1000000 0.0000000 0.1000000
#> D 0.0000000 0.0000000 0.1000000 0.2666667 0.0000000 0.1666667
#> E 0.1333333 0.1000000 0.0000000 0.0000000 0.2333333 0.0000000
#> F 0.0000000 0.0000000 0.1000000 0.1666667 0.0000000 0.2666667
#> attr(,"class")
#> [1] "cohesion_matrix" "matrix"          "array"
\end{Soutput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
df |>
  dist() |>
  cohesion_matrix() -> cohesion
\end{Sinput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
community_clusters(cohesion)
\end{Sinput}
\begin{Soutput}
#>   point cluster
#> A     A       1
#> B     B       1
#> C     C       2
#> D     D       3
#> E     E       1
#> F     F       3
\end{Soutput}
\end{Schunk}

In this example, three clusters are identified with these six points.
Points \texttt{A}, \texttt{B}, and \texttt{E} fall into cluster 1. Point
\texttt{C} is in cluster 2 and points \texttt{D} and \texttt{F} fall
into cluster 3.

@@

\begin{Schunk}
\begin{Sinput}
local_depths(cohesion)
\end{Sinput}
\begin{Soutput}
#>         A         B         C         D         E         F 
#> 0.5666667 0.4666667 0.4333333 0.5333333 0.4666667 0.5333333
\end{Soutput}
\end{Schunk}

In this case, the deepest point is \texttt{A}.

@@

\begin{Schunk}
\begin{Sinput}
strong_threshold(cohesion)
\end{Sinput}
\begin{Soutput}
#> [1] 0.1222222
\end{Soutput}
\end{Schunk}

In this case, the threshold is \texttt{0.12}.

@@

\begin{Schunk}
\begin{Sinput}
cohesion_strong(cohesion)
\end{Sinput}
\begin{Soutput}
#>           A         B         C         D         E         F
#> A 0.2333333 0.1333333 0.0000000 0.0000000 0.1333333 0.0000000
#> B 0.1333333 0.2333333 0.0000000 0.0000000 0.0000000 0.0000000
#> C 0.0000000 0.0000000 0.2333333 0.0000000 0.0000000 0.0000000
#> D 0.0000000 0.0000000 0.0000000 0.2666667 0.0000000 0.1666667
#> E 0.1333333 0.0000000 0.0000000 0.0000000 0.2333333 0.0000000
#> F 0.0000000 0.0000000 0.0000000 0.1666667 0.0000000 0.2666667
#> attr(,"class")
#> [1] "cohesion_matrix" "matrix"          "array"
\end{Soutput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
graphs <- community_graphs(cohesion)
graphs[["G_strong"]]
\end{Sinput}
\begin{Soutput}
#> IGRAPH dcdcffa UNW- 6 3 -- 
#> + attr: name (v/c), weight (e/n)
#> + edges from dcdcffa (vertex names):
#> [1] A--B A--E D--F
\end{Soutput}
\end{Schunk}

Here we see that there are three connected components, points \texttt{A}
and \texttt{B}, and \texttt{A} and \texttt{E} which form the first
cluster, and points \texttt{D} and \texttt{F} which form another.

@@

The final category of function is functions for data visualization. We
can begin by visualizing the points in data frame \texttt{df} (Figure
\ref{fig:fig1}). When visualizing these points, it is important to have
a square plot so as to not distort the distances. When using the
\CRANpkg{ggplot2} package for this visualization, you can use the
\texttt{aspect.ratio\ \ =\ 1} argument in the \texttt{theme()} function.
If using the \texttt{plot()} function included in the base library, you
can set the plot area to be a square by using \texttt{par(pty\ =\ "s")}
prior to calling the \texttt{plot()} function, as shown below.

\begin{Schunk}
\begin{Sinput}
library(ggplot2)
ggplot(df, aes(x1, x2)) +
  geom_text(label = rownames(df)) + 
  theme(aspect.ratio = 1)
\end{Sinput}
\begin{figure}
\includegraphics{manuscript_files/figure-latex/fig1-1} \caption[Visualize the points from data frame `df`]{Visualize the points from data frame `df`}\label{fig:fig1}
\end{figure}
\end{Schunk}

We can pass the cohesion matrix to the
\texttt{plot\_community\_graphs()} function to view the relationship
between points (Figure \ref{fig:fig2}).

\begin{Schunk}
\begin{Sinput}
plot_community_graphs(cohesion)
\end{Sinput}
\begin{figure}
\includegraphics{manuscript_files/figure-latex/fig2-1} \caption[PaLD graph displaying the relationship between the points in data frame `df`]{PaLD graph displaying the relationship between the points in data frame `df`}\label{fig:fig2}
\end{figure}
\end{Schunk}

The \texttt{layout} argument allows the user to pass a matrix to dictate
the layout of the graph. For example, if we wanted the graph to match
the visualization displayed in Figure \ref{fig:fig1}, we can pass
\texttt{as.matrix(df)}, or a matrix of the data frame \texttt{df} to the
\texttt{layout} argument (Figure \ref{fig:fig3}. Notice in the code
below, we set \texttt{par(pty\ =\ "s")} in order to ensure that we have
a square plotting region. Additionally, this
\texttt{plot\_community\_graphs()} function will also permit parameters
that can be passed to the \texttt{plot.igraph()} function. For example,
we can pass arguments to the \texttt{plot.igraph} function via the
\texttt{...}, for example to increase the vertex size and change the
vertex label color, we can specify \texttt{vertex.size\ =\ 15} and
\texttt{vertex.label.color\ =\ "white"}.

\begin{Schunk}
\begin{Sinput}
par(pty = "s")

plot_community_graphs(cohesion, 
                      layout = as.matrix(df),
                      vertex.size = 15,
                      vertex.label.color = "white")
\end{Sinput}
\begin{figure}
\includegraphics{manuscript_files/figure-latex/fig3-1} \caption[PaLD graph displaying the relationship between the points in data frame `d`, matching the original layout in Figure 1]{PaLD graph displaying the relationship between the points in data frame `d`, matching the original layout in Figure 1}\label{fig:fig3}
\end{figure}
\end{Schunk}

@@

The first example if from a subset of tissue gene expression data from
\citet{zilliox2007gene}, \citet{mccall2011gene}, and
\citet{mccall2014gene}, obtained from the \textbf{tissuesGeneExpression}
bioconductor package \citep{tissue}. A \texttt{dist} object was created
using this data set and is included the \CRANpkg{pald} package in an
object called \texttt{tissue\_dist}.

The \texttt{tissue\_dist} object is a \texttt{dist} object resulting in
a distance matrix with 189 rows and 189 columns.

We can create the cohesion matrix using the \texttt{cohesion\_matrix}
function.

\begin{Schunk}
\begin{Sinput}
tissue_cohesion <- cohesion_matrix(tissue_dist)
\end{Sinput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
community_clusters(tissue_cohesion) |>
  dplyr::count(cluster, point)
\end{Sinput}
\begin{Soutput}
#>    cluster       point  n
#> 1        1 endometrium 15
#> 2        1      kidney 39
#> 3        2 hippocampus 31
#> 4        3  cerebellum 26
#> 5        4  cerebellum  1
#> 6        5       colon 33
#> 7        6       colon  1
#> 8        7       liver  7
#> 9        8  cerebellum  1
#> 10       9       liver 17
#> 11      10  cerebellum  2
#> 12      11       liver  2
#> 13      12  cerebellum  1
#> 14      13  cerebellum  4
#> 15      14  cerebellum  2
#> 16      15  cerebellum  1
#> 17      16    placenta  2
#> 18      17    placenta  1
#> 19      18    placenta  3
\end{Soutput}
\end{Schunk}

@@

We can also display the relationships between tissue samples using the
\texttt{plot\_community\_graphs()} function (Figure \ref{fig:fig4}). For
clarity of the display, we show how to remove the labels using
\texttt{show\_labels\ =\ FALSE}. We will instead color by the labels by
passing these to the \texttt{vertex.color} parameter through the
\texttt{...} to the \texttt{plot.igraph} function. Similarly, we can add
a legend using the \texttt{legend()} function, as you would for a
\CRANpkg{igraph} visualization. Additionally, we use the
\texttt{edge\_width\_factor} and \texttt{emph\_strong} arguments to
adjust the width of the lines between and within PaLD clusters.

\begin{Schunk}
\begin{Sinput}
labels <- rownames(tissue_cohesion)
plot_community_graphs(tissue_cohesion,
                      show_labels = FALSE,
                      vertex.size = 4,
                      vertex.color = as.factor(labels),
                      edge_width_factor = 35,
                      emph_strong = 5) 
legend("topleft", 
       legend = unique(as.factor(labels)), 
       pt.bg = unique(as.factor(labels)),
       col = "black",
       pch = 21)
\end{Sinput}
\end{Schunk}

\begin{Schunk}
\begin{figure}
\includegraphics[width=1\linewidth]{fig5} \caption[PaLD clustering of tissue data]{PaLD clustering of tissue data. The line colors indicate the PaLD clusters, the point colors indicate the tissue classification.}\label{fig:fig4}
\end{figure}
\end{Schunk}

\hypertarget{cognate-based-language-families}{%
\subsection{Cognate-based Language
Families}\label{cognate-based-language-families}}

This example performs a PaLD analysis on a data set from \citet{dyen92}
that examines the relationship between 87 Indo-European languages from
the perspective of cognates. A \texttt{dist} object was created from
this data set and included in the \CRANpkg{pald} package in an object
called \texttt{cognate\_dist}.

This example will demonstrate how you can apply functions in the
\CRANpkg{igraph} package to objects output from the \CRANpkg{pald}
package. We can first use the \texttt{cohesion\_matrix()} function to
calculate the cohesion matrix and the \texttt{community\_graphs()}
function to create a list with the weighted community graph, the
weighted community graph with only strong ties included (and all others
set to 0), and the layout. From this, we can extract the graph with only
the strong ties, here called \texttt{cognate\_graph\_strong}.

\begin{Schunk}
\begin{Sinput}
cognate_cohesion <- cohesion_matrix(cognate_dist)
cognate_graphs <- community_graphs(cognate_cohesion)

cognate_graph_strong <- cognate_graphs[["G_strong"]]
\end{Sinput}
\end{Schunk}

We can then use the \texttt{neighbors()} function from the
\CRANpkg{igraph} package to extract the neighbors in this graph. For
example, if we wanted to extract all neighbors where the language is
``French'', we would run the following.

\begin{Schunk}
\begin{Sinput}
french_neighbors <- igraph::neighbors(cognate_graph_strong, "French")
french_neighbors
\end{Sinput}
\begin{Soutput}
#> + 8/87 vertices, named, from 64a847b:
#> [1] Italian         Ladin           Provencal       Walloon        
#> [5] French_Creole_C French_Creole_D Spanish         Catalan
\end{Soutput}
\end{Schunk}

Similarly, we can print the associated neighborhood weights by
subsetting the cohesion matrix.

\begin{Schunk}
\begin{Sinput}
cognate_cohesion["French", french_neighbors]
\end{Sinput}
\begin{Soutput}
#>         Italian           Ladin       Provencal         Walloon French_Creole_C 
#>      0.01997696      0.02094596      0.02871174      0.03258771      0.02406057 
#> French_Creole_D         Spanish         Catalan 
#>      0.02406057      0.01679733      0.01859688
\end{Soutput}
\end{Schunk}

@@

Here, we will demonstrate how to use \texttt{exdata3}. These points were
generated from bivariate normal distributions with varying means and
variances. There are eight ``true'' clusters.

We will demonstrate how we can compare PaLD to two clustering methods:
\emph{k}-means and hierarchical clustering. The code below calculates
the cohesion matrix (\texttt{exdata\_cohesion}) as well as the clusters
via PaLD (\texttt{exdata\_pald}), \emph{k}-means
(\texttt{exdata\_kmeans}) and hierarchical clustering
(\texttt{exdata\_hclust}).

\begin{Schunk}
\begin{Sinput}
exdata_cohesion <- exdata3 |>
  dist() |>
  cohesion_matrix()

exdata_pald <- community_clusters(exdata_cohesion)$cluster

exdata_kmeans <- kmeans(exdata3, 8)$cluster

exdata_hclust <- exdata3 |>
  dist() |>
  hclust() |>
  cutree(k = 8) 
\end{Sinput}
\end{Schunk}

We can compare this to the clustering generated by \emph{k}-means and
hierarchical clustering (Figure \ref{fig:fig5}).

\begin{Schunk}
\begin{Sinput}
par(mfrow = c(1, 3), pty = "s")
plot(
  exdata3,
  pch = 16,
  col = pald_colors[exdata_pald],
  xlab = "",
  ylab = "",
  main = "PaLD Clusters"
)
plot(
  exdata3,
  pch = 16,
  col = pald_colors[exdata_kmeans],
  xlab = "",
  ylab = "",
  main = "K-Means Clusters (k = 8)"
)
plot(
  exdata3,
  pch = 16,
  col = pald_colors[exdata_hclust],
  xlab = "",
  ylab = "",
  main = "Hiearchical Clusters (k = 8)"
)
\end{Sinput}
\begin{figure}
\includegraphics{manuscript_files/figure-latex/fig5-1} \caption[PaLD clustering of randomly generated example data (from Figure 4D from Berenhaut et al]{PaLD clustering of randomly generated example data (from Figure 4D from Berenhaut et al. (2022)) compared to *k*-means and hierarchical clustering with k = 8.}\label{fig:fig5}
\end{figure}
\end{Schunk}

Cohesion is particularly useful when considering data with varying local
density, see discussion in \citep{berenhaut2022social}. Note that the
PaLD algorithm is able to detect the eight natural groups within the
data without the use of any additional inputs (e.g., number of clusters)
nor optimization criteria. Despite providing the ``correct'' number of
clusters (i.e., \(k = 8\)) both \emph{k}-means and hierarchical
clustering did not give the desired result.
